{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}

### 2017-03-01 proposal/thesprop.tex
thesprop: split pintos and tm into their own chapters

		This work is {\bf completed} and was published at OOPSLA 2016.
	\item {\bf Educational use (Chapter \ref{chap:410}).}
		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
		%, investigating the following questions in particular:
		%\begin{itemize}
		%	\item Does Landslide improve project submission quality when used as a debugging aid?
		%	\item Does Landslide teach students lessons in concurrent programming
		%\end{itemize}

		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them and follow them up with an evaluation.
	\item {\bf Broader impact (Chapter \ref{chap:lipservice}).}
		Landslide includes several new techniques for testing kernels and thread libraries with little or no manual instrumentation required.
		So far, a fully-automatic testing mode is available only for 15-410 thread library projects.
		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}. %and ``real-world'' Linux programs.
		%I will conduct a user study in which students at another university test their Pintoses with Landslide,
		%and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
		I will also extend Landslide's concurrency model to support Hardware Transactional Memory (HTM), a new concurrent programming technique \cite{transactional-memory}, and test several ``real-world'' HTM programs.

		This will be {\bf mostly new work}, as Landslide presently supports testing Pintoses only with some manual effort,
		%and does not yet feature instrumentation to run on Linux programs.
		and will need its concurrency model extended to incorporate the HTM programming model.
\end{itemize}

@@

\input{overview}
\input{quicksand}
\input{410}
\input{lipservice}
\input{related}

@@

\begin{itemize}
	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{sec:pintos-todo}).
	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{sec:pintos-todo}).
	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
	\item 2017 October-November: Deploy 410 TA grading study again for more data.
	\item 2017 December: Begin writing thesis.
	\item 2018 January-April: Write thesis. Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
	\item 2018 April-May: Defend \& graduate.
\end{itemize}

>>>>

		This work is {\bf completed} and was published at OOPSLA 2016.
	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
		%, investigating the following questions in particular:
		%\begin{itemize}
		%	\item Does Landslide improve project submission quality when used as a debugging aid?
		%	\item Does Landslide teach students lessons in concurrent programming
		%\end{itemize}

		So far, the fully-automatic testing mode is available only for 15-410 thread library projects.
		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
		I will then collaborate with those schools to deploy Landslide to their operating systems students.

		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them
		and extend the existing evaluation,
		as well as deploy a Pintos port.
	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
		Transactional Memory (TM) is a relatively new concurrent programming technique \cite{transactional-memory}
		which is not yet addressed by modern model checkers.
		I will extend Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
		and test several ``real-world'' TM programs.

		This will be {\bf new work},
		involving some implementation effort, some proofs, and designing a new evaluation.
		%, as Landslide presently supports testing Pintoses only with some manual effort,
		%and does not yet feature instrumentation to run on Linux programs.
		%and will need its concurrency model extended to incorporate the HTM programming model.
\end{itemize}

@@

\input{overview}
\input{quicksand}
\input{410}
\input{pintos}
\input{tm}
\input{related}

@@

\begin{itemize}
	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
	\item 2017 October-November: Deploy 410 TA grading study again for more data.
	\item 2017 December: Begin writing thesis.
	\item 2018 January-April: Write thesis. Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
	\item 2018 April-May: Defend \& graduate.
\end{itemize}

### 2017-03-01 proposal/tm.tex
thesprop: split pintos and tm into their own chapters

\chapter{Broader Impact}
\label{chap:lipservice}

In this chapter I propose to add several enhancements to Landslide to apply its techniques beyond CMU's walls.
%with little to no manual instrumentation effort required.
This is the third of the three projects I am proposing for this thesis and so far exists only in dreams.
The project will have two parts: adding support for Pintos \cite{pintos}, and adding support for Transactional Memory \cite{transactional-memory}.

\section{Pintos}
\label{sec:pintos-todo}

The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos}, provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
However, the existing scheduler and concurrency primitives which Pintos provides somewhat limits the remaining concurrency-critical code left to write,
hence limiting the degree to which Landslide can test student submissions.
This is in contrast to the Pebbles assignments at CMU, in which students must implement all the functionality listed in Section~\ref{sec:overview-pintos}.
On the other hand, the upside of Pintos providing substantially more basecode than Pebbles is that most, if not all, of Landslide's kernel instrumentation can be done automatically,
using the names of the core scheduler functions already provided.
For the experiments in Chapter~\ref{chap:quicksand}, I already implemented rudimentary support to test Pintos kernels, but more remains to be done.

{\bf Stuff I already did.}
The existing Pintos support includes
a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
handling several quirks of the basecode's scheduling behaviour (ask me if you really care to know);
extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
as well as the fact that kernel memory is not direct-mapped like in Pebbles;
new thread-liveness code to detect when a test case terminates;
skipping some busy-wait loops in the boot sequence used for device communication,
a complicated fall-back case in Landslide's timer injection algorithm the workings of which I don't even remember anymore,
and extending the lock-set analysis to include Pintos's interrupt-disabling infrastructure (which was actually pretty tricky when using vector clocks for pure-HB).
All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.

{\bf Stuff I still need to do.}
However, as you might have guessed, there are still a few cases where the existing instrumentation is not fully general.
When preparing Chapter~\ref{chap:quicksand}'s experiments, I needed to adjust some student submissions to be compatible with Landslide's existing annotation process.
In order to distribute Landslide for student use beyond CMU, it will need to support these cases automatically.
These include:
\begin{itemize}
	\item Allowing for the priority scheduler's runqueues to be implemented as an array of queues, rather than the single queue head as provided in the basecode.
	\item Insert the {\tt tell\_landslide\_sleeping} annotation more intelligently than looking for a {\tt list\_insert\_ordered} call (a basecode function which most, but not all, students use).
	\item Some versions of the basecode distribute a {\tt sema\_up} implementation which {\tt yield}s unconditionally, which Landslide must bypass.
	\item Some students use {\tt timer\_sleep} or {\tt while (!flag) continue} when they should use {\tt yield}, which can lead to false-positive deadlocks if not automatically replaced.
	\item I'm sure I've forgotten some less-common ones, which I expect to handle on-the-fly when students email me Landslide crash reports.
\end{itemize}
When fixing my sample of Pintoses by hand, there were also myriad deterministic bugs, such as use-after-frees e.g.~arising from unsafe {\tt strlen} calls.
As with Pebbles student experiments, however, I intend the students to encounter such bug reports and fix them on their own, despite not being concurrency bugs.
It would be a Ph.D. unto itself to automatically ensure stable determinized execution of arbitrary student code.
Actually, probably outright impossible.

{\bf New emulation platform.}
Most importantly, I will need to free Landslide from its dependence on Simics, which requires paid licenses for use beyond CMU's walls.
Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
I am presently studying which of these will most suit Landslide's needs,
but so far Bochs seems like the best bet, which happens to be the simulator used at Berkeley, Stanford, and U. of Chicago anyway.

{\bf Experimental goals.}
Having only one semester (ideally; or two, if my timeline ends up delayed) to test Landslide with Pintos students,
I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
This will perhaps shed light on the advantages and/or shortcomings of either project,
leading to recommendations for improving either project's educational value (independently of using MC).

\section{Transactional Memory}
\label{sec:tm}

Transactional Memory (TM) is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
A transactional program surrounds its critical section(s) with transaction begin/end statements, which ensure that no other thread can observe an intermediate state during the transaction.
If a conflict is observed, the transaction {\em aborts}, rolling the program back to the transaction's initial state, and executing an optional back-up code path.
The programmer may also explicitly abort the transaction using an abort statement.
An example transactional program is shown in Figure~\ref{fig:txn-example}.

@@

\subsection{HTM}

@@

\subsection{STM}

@@

\subsection{Hybrid HTM/STM}

@@

>>>>

\chapter{Transactional Memory}
\label{chap:tm}

In this chapter I propose to extend Landslide's concurrency model to support Transactional Memory (TM) \cite{transactional-memory}.
This is the last of the four projects I am proposing for this thesis and so far exists only in dreams.

TM is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
A transactional program surrounds its critical section(s) with transaction begin/end statements, which ensure that no other thread can observe an intermediate state during the transaction.
If a conflict is observed, the transaction {\em aborts}, rolling the program back to the transaction's initial state, and executing an optional back-up code path.
The programmer may also explicitly abort the transaction using an abort statement.
An example transactional program is shown in Figure~\ref{fig:txn-example}.

@@

\section{HTM}

@@

\section{STM}

@@

\section{Hybrid HTM/STM}

@@

\section{Evaluation Plan}

% TODO
